---
title: "Manuscript_Analyses"
author: "Jessica McClintock"
date: "`r Sys.Date()`"
output: pdf_document
---

## Setup

```{R Setup}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(BatchQC)
library(SummarizedExperiment)
```

### Data Retrieval
All data is hosted on github. Please see the following for information on how to
retrieve each data set used in the analysis.

#### App Comparison Data
Excel sheet with e a table comparing the different R shiny applications
available for batch effect analysis.

```{R App Comparison Data}
app_comparison_location <- paste0("https://raw.githubusercontent.com/jessmcc22",
    "/lab_meeting/refs/heads/devel/May_15_25/Data",
    "/ShinyAppBatchCorrectionToolComparison.csv")

app_comparison_data <- read.csv(app_comparison_location, header = TRUE)
```

#### ADNI Data
Alzheimer's imaging data provided by Taki Shinohara (UPenn). In this data
set, three different scanners were used (Siemens Healthineers (n=96), Philips
Medical Systems (n = 96) or GE Healthcare (n = 287)) and in a previous analysis
done with DeepComBat, a batch effect was seen between Siemens and Philips/GE
(combined as one batch) scanners. We are hoping to recreate this batch effect
with BatchQC when comparing Alzheimer diagnosis (cognitively normal (CN), late
mild cognitive impairment (LMCI) and Alzheimer disease (AD)). To note, this data
includes up to 6 images for each subject taken at different visits, thus there
are 2515 unique images.

```{R ADNI data}
ADNI_data_location <- paste0("https://raw.githubusercontent.com/ntustison",
    "/CrossLong/refs/heads/master/Data/reconciled_ANTsSST.csv")

ADNI_additional_metadata_location <- paste0("https://raw.githubusercontent.com",
    "/jessmcc22/lab_meeting/refs/heads/devel/May_15_25/Data",
    "/protocol_preprocessed_2019_03_01.csv")

ADNI_data <- read.csv(ADNI_data_location, header = TRUE)
ADNI_additional_metadata <- read.csv(ADNI_additional_metadata_location,
    header = TRUE)
```

## Introduction

### Figure 1: R Shiny Batch Effect Application Comparison
Many different R Shiny apps exist that streamline data exploration, analysis and
visualization of batch correction analyses. The methods and tools available in
some common applications are compared in the below table. The information in
this table was obtained from the accompanying manuscript publication for each
application. The table presented in our paper was formatted in excel and a
simpler, unformatted table with the same information is presented below.

```{R Application Comparison, results = "asis"}
knitr::kable(app_comparison_data)
```

## Methods

### Lambda Statistic

### Model Selection Checks

### Data Examples

#### ADNI Data

The original data set includes multiple images for each subject taken in a
longitudinal manner at multiple different visits. For our analysis, we are
interested in comparing the baseline images at the first visit. The set of
images also contain images of various locations in the brain and for this
analysis, we will focus on the temporal location.

```{R Analysis Options}
VISIT_NUMBER <- 0
BRAIN_LOCATION <- "temporal"
```

Tidy the downloaded file into metadata and cortical thickness dataframes

```{r Create Thickness and Metadate DF}
ADNI_data_t0 <- ADNI_data[order(ADNI_data$IMAGE_ID),] %>% 
    filter(VISIT == VISIT_NUMBER)
Image_ID <- as.character(ADNI_data_t0$IMAGE_ID)
ADNI_metadata <- as.data.frame(ADNI_data_t0[,1:8], row.names = Image_ID)
cortical_thickness <- (as.data.frame(ADNI_data_t0[,9:70], row.names = Image_ID))
```

Subset cortical thickness measures based on location of interest.

```{r Create Location Assays}
specific_location_data <- cortical_thickness %>%
    select(contains(BRAIN_LOCATION)) %>% 
    t()
cortical_thickness <- t(cortical_thickness)

```

Tidy the data by creating a column name that contains all the scan locations and
a column name with the thickness values

```{r Tidy the Data}
tidy_ADNI <- ADNI_data_t0 %>%
    pivot_longer(9:dim(ADNI_data_t0)[2],
        names_to = "LOCATION",
        values_to = "THICKNESS")
```

The acquisition site is a three digit code within the `ID` column prior to the
`_S_` identifier. Use this to create a new column containing only the
acquisition site number to be used as a batch variable.

```{r Create Site Variable}
ADNI_metadata <- ADNI_metadata %>% 
    separate_wider_position(ID, 
        c(SITE = 3, 1, SAMPLE_ID = 6),
        cols_remove = FALSE)

ADNI_metadata <- as.data.frame(ADNI_metadata)
```

The scanner data is contained in the extra metadata file (not hosted on github,
provided by Taki to Jessica). Using the ImageUID in the scanner data, combine
this with the ImageID in the metadata and retain the scanner type
(Manufacturer). Modify manufacturer to exclude spaces.

```{R Obtain Scanner Info}
manufacturer_data <- ADNI_additional_metadata %>%
    select(imageUID, Manufacturer)
metadata_scanner <- merge(ADNI_metadata, manufacturer_data,
    by.x = "IMAGE_ID",
    by.y = "imageUID",
    all.x = TRUE)

for (i in 1:length(metadata_scanner$Manufacturer)){
    if(metadata_scanner$Manufacturer[i] == "GE MEDICAL SYSTEMS" ||
            metadata_scanner$Manufacturer[i] == "Philips Medical Systems"){
        metadata_scanner$Manufacturer[i] <- "GE_PHILLIPS"
    }
}

row.names(metadata_scanner) <- metadata_scanner$IMAGE_ID
```

Convert the columns to factors for downstream analyses.

```{R Convert to Factors}
metadata_scanner$SITE <- as.factor(metadata_scanner$SITE)
metadata_scanner$VISIT <- as.factor(metadata_scanner$VISIT)
metadata_scanner$DIAGNOSIS <- as.factor(metadata_scanner$DIAGNOSIS)
metadata_scanner$SEX <- as.factor(metadata_scanner$SEX)
metadata_scanner$Manufacturer <- as.factor(metadata_scanner$Manufacturer)
```

Create a summarized experiment object.

```{R SE Object Creation}

ADNI_se <- SummarizedExperiment(
    assays = list(cortical_thickness = specific_location_data),
    colData = metadata_scanner,
    metadata = list(metadata = metadata_scanner))
```

## Results

### Simulation Data

### ADNI Data

#### Evidence of Batch Effect
Using BatchQC, we can view the raw data and identify evidence of batch effect
within the data. 

Below, we set up the variables that we are interested in analyzing: the raw data
assay, the Manufacturer machine is the variable we suspect a batch effect to
reside, and the diagnosis variable is where we would expect true differences
between samples to be found.

```{R Batch Effect Analysis Variables}
cortical_assay <- "cortical_thickness"
batch <- "Manufacturer"
covar <- "DIAGNOSIS"
```

We plot box plots of the percent variation explained by our batch and variable
of interest. Our first plot shows individual variation and the second plot shows
type 2 variation. We see that the Manufacturer (batch effect) explains a similar
amount of variation as Diagnosis (experimental variable), thus indicating the
presence of a batch effect.

```{R Variation Analysis}
ex_variation <- BatchQC::batchqc_explained_variation(se = ADNI_se,
    batch = batch,
    condition = covar,
    assay_name = cortical_assay)
# individual variation
EV_boxplot <- BatchQC::EV_plotter(batchqc_ev = ex_variation[[1]])
EV_boxplot

# Type 2 variation
EV_boxplot_type_2 <- BatchQC::EV_plotter(batchqc_ev = ex_variation[[2]])
EV_boxplot_type_2

```

We plot a PCA plot of the raw data where color represents batch and shape
represents the diagnosis. The ellipses shows the batch groupings and the
distinct ellipses indicate that there is a batch effect within the data.

```{R ADNI PCA}
pca_plot <- BatchQC::PCA_plotter(se = ADNI_se, 
    nfeature = 8, color = batch,
    shape = covar, batch = batch,
    assays = cortical_assay,
    xaxisPC = 1, yaxisPC = 4, log_option = FALSE)
pca_plot$plot
```

#### Lambda Statistic to Confirm need for Batch Correction
We can confirm the need for a batch correction using our new tools. First, we
will determine if our experiment is a balanced or unbalanced design.

```{R Check for Balanced Design}

```

Next, we calculate the lambda statistic

```{R lambda statistic}

```

Based on this result...

#### Check Model Fit
We will use our AIC model to check the fit of the data.

```{R Model Fit}

```

Since...., we can proceed with....

#### ComBat Batch Correction
Use BatchQC to correct for batch effect using ComBat.

```{R ComBat Correction}
ADNI_se <- BatchQC::batch_correct(se = ADNI_se, 
    method = "ComBat",
    assay_to_normalize = cortical_assay, 
    batch = batch,
    covar = c(covar, "AGE", "SEX"),
    output_assay_name = "ComBat_cortical_thickness")

ComBat_assay <- "ComBat_cortical_thickness"
```

### Other Data

```{R Data Preprocessing}

```

